Improving reproducibility of scientific software using Nix/NixOS: A case study on the preCICE ecosystem

### 2024-05-10 paper/paper.tex
drop(paper): Example precice nix paths
chore: drop comparison

All contents, including source files and resulting build artifacts, are stored inside the Nix store.
Per default, the Nix store resides in the path \texttt{/nix/store} on the file system.
The naming scheme with the preCICE 2.5.0 package as an example is

\begin{itemize}
  \item \texttt{/nix/store/0a5gw3l\ldots-precice-2.5.0}
  \item \texttt{/nix/store/9gpyg37\ldots-precice-2.5.0}
\end{itemize}

This eases checking the Nix store for integrity and ensures that builds using the same inputs are performed only once.

@@

\section{Comparison}

This chapter compares Nix with two widely used solution in HPC environments, Spack and EasyBuild.
There was a similar comparison done at FOSDEM 2018~\footnote{\url{https://archive.fosdem.org/2018/schedule/event/installing_software_for_scientists/}} that also included conda and Guix, but this was by the release manager of EasyBuild and we want to provide an update with this comparison.
The comparison first takes a high level look on some statistics and how they compare and after that we will discuss, usability, reproducibly and performance in more detail.

\subsection{Overview}

The first table~\ref{table:comparison-overview} shows some basic stats about the amount of packages the solutions provide out of the box and how they are implemented.
This is important because the higher the number the more likely it is that your package is already packaged and does not require additional work.
It is also important to look how packages are implemented because, if a package needs to be packaged by for example researches, domain specific languages have the downside that, researches now have to get familiar with this DSL, while general purpose languages such as python do not have this downside because they are already used in their field.
Although this is a downside, DSL usually are design with a purpose in mind and can get you faster to the desired outcome.

\begin{figure*}
  \normalsize
  \caption{Overview of package count and how packages are implemented}
  \label{table:comparison-overview}
  \centering
  \begin{tabular}{|p{2cm}|p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    \textbf{Metrics} & \textbf{Nix} & \textbf{Spack} & \textbf{EasyBuild} \\ \hline
    Packages~\footnote{\label{foot:repology}\url{https://repology.org/repositories/statistics/total}} & 91800 & 7910 & 19500 \\ \hline
    Unique packages & 73200 & 6610 & 3470 \\ \hline
    Package implementation & Nix Domain Specific Language & Python & Python \\ \hline
  \end{tabular}
\end{figure*}

The installing and updating process is documented in the second table~\ref{table:comparison-installation} and differs widely, from custom install scripts over to using pip or just using git clone.

Spack's way of installing has some problems, with using \texttt{git clone} on the latest default branch, in their case \texttt{develop}, this has the implication, that you are not using the latest stable release of the software but rather a random commit, that might have broken something recently.
It also has the problem that you cannot easily update the package repository independent from the package manager, this requires additional work from the users if they require this.

It is also worth mentioning that all 3 solutions do not require root for installation but Nix has the option for a system wide installation of the Nix daemon, that can then be used to install packages locally into the users PATH, without requiring root permissions.

\begin{figure*}
  \normalsize
  \caption{How the package manager are being installed and how are they updated}
  \label{table:comparison-installation}
  \centering
  \begin{tabular}{|p{2cm}|p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    \textbf{Metrics} & \textbf{Nix} & \textbf{Spack} & \textbf{EasyBuild} \\ \hline
    Installation way & install script~\footnote{\url{https://nixos.org/download.html\#download-nix}} & \texttt{git clone}~\footnote{\url{https://spack.readthedocs.io/en/latest/getting_started.html\#installation}} & \texttt{pip install} or afterwards self install using \texttt{eb} executable~\footnote{\url{https://docs.easybuild.io/installation/}} \\ \hline
    Installation mode & single user/multi user & single user & single user \\ \hline
    Updating self & \texttt{nix-channel} / nix flake for pinning & \texttt{git pull} & using the \texttt{eb} executable \\ \hline
  \end{tabular}
\end{figure*}

The basic usage of all three tools is pretty similar and they support searching and installing package in a simple manner, all this is show in table~\ref{table:comparison-usage}.
The difference in installing software comes down to the unique way that Nix installs software, it downloads/builds the paths into its local store, that can either be located in the user's home directory or in the path \texttt{/nix/store} depending on the installation.
\texttt{nix-shell} later only puts these downloaded paths into the \texttt{\$PATH} environment variable.
Spack and EasyBuild on the other hand use the widely used \texttt{module} system that usually is already found on HPC clusters, it allows loading software into the current environment.

Installing old software also differs for all three solutions, Nix on the one hand keeps some old, widely used software packages around and maintains those, examples are \texttt{boost} that is available in different versions or e.g. \texttt{hdf5} that is available in the latest version and version \texttt{1.10}.
The other way of installing older software on nix is using something that is called \texttt{overrideAttrs}, it can be used to change anything of a package, such as the used sources, meaning it can be used to replace it with an older/newer version or it can be used to add an patch, add a new dependency or install additional files into the \texttt{\$out} directory.
An example would be the following \texttt{nix-shell -p 'precice.overrideAttrs(\_: \{ version = "2.4.0"; src = \ldots \} )'}, it replaces the preCICE version with 2.4.0 and adds the old source, that will then trigger a new build.
Spack and EasyBuild on the other hand, keep the old software versions in the specifications or keep the whole specification, so you can reinstall an old version by specifying the available version, for example, \texttt{spack install gcc@6.4.0}
If the required version is not available, you have to edit the current package definition and add the required version.

To achieve reproducibility, it is necessary that we can install software from a previous revision of the packages.
Nix allows us to simply specify the channel revision we want to use to build our software, it then evaluates the requested packages and builds it.
This can easily be done without affecting the current system by specifying it in the shell command \texttt{nix-shell -p pan -I nixpkgs=pathToRevision/urlToTarball}, it now only uses this channel for building the requested package.
Spack, can achieve something similar by checking out your Spack version at a previous revision and then installing the package requested package, but this approach has a downside that it affects the whole Spack installation, and now also work with an older Spack tool.
On the other hand, EasyBuild, could be installed using pip at an earlier version, but it also offers something different called toolchains, we will look at shortly.

Next, we look at how you can build a package with a specific compiler version using specific compiler flags.
The flexibility of Nix overrides allows different levels of rebuilds of packages with a specific compiler, you can either just rebuild the requested package with specific compiler, this for example rebuilds curl with clang8 \texttt{nix-shell -p '(curl.override \{ stdenv = clang8Stdenv;\})'} or you can rebuild the package and all required dependencies with a specific compiler by overriding the \texttt{stdenv} on nixpkgs level, \texttt{nix-build -E 'with (import <nixpkgs> \{ config.replaceStdenv = \{ pkgs \}: pkgs.clang8Stdenv; \}); curl'}.
This is possible because every c/c++/\ldots package, is build using the \texttt{stdenv} that defines the compiler and default compile flags.
While Nix provides a default stdenv, that is used to compile all packages, it is possible to override and replace the package with your another stdenv provided by nixpkgs, such as the clang8Stdenv, you can write your own stdenv, or you can extend the current stdenv with flags.
By default, there are some useful helpers provided, such as \texttt{makeStaticLibraries} that allows compiling any library as a static library, or \texttt{enableDebugging} that sets the compiler flags \texttt{-ggdb -Og} and can then be enabled on the package plus all dependencies.
Spack provides an easy way to specify the compiler using \texttt{spack install precice \%clang} and only requires a registered compiler, it can be a system compiler, or the compiler can be installed using Spack.
On this note, Spack tooling and interface to override compilers, flags and more is way more user friendly and readable than using and invoking the Nix DSL within a one line command call.
EasyBuild has a whole different system, where packages have defined toolchains.\footnote{\url{https://docs.easybuild.io/common-toolchains/}}
A toolchain is a set of packages defined at a specific version and users can create their own toolchains.
For example, EasyBuild provides a foss toolchain that contains packages such as, binutils, gcc, OpenMPI, openblas and more and packages are provided against a specific toolchain version such as foss-2022b.
Packages are then provided with toolchains in mind and are never deleted, so installing a package with a different compiler can be achieved by selecting a different eb script with a different toolchain version.
On the other hand this results in a less elegant user flow, because if a user wants to install a package with a toolchain that is not provided they have to fork a package definition and update the toolchain.\\

\begin{figure*}
  \normalsize
  \caption{Usage of the package manager}
  \label{table:comparison-usage}
  \centering
  \begin{tabular}{|p{2cm}|p{3.5cm}|p{3.5cm}|p{3.5cm}|}
    \hline
    \textbf{Metrics} & \textbf{Nix} & \textbf{Spack} & \textbf{EasyBuild} \\ \hline
    Search & \texttt{nix-env -qa 'mpi'}~\footnote{\url{https://nix.dev/tutorials/first-steps/}} & \texttt{spack list 'query'}~\footnote{\url{https://spack.readthedocs.io/en/latest/basic_usage.html}} & \texttt{eb -search mpi}~\footnote{\url{https://spack.readthedocs.io/en/latest/basic_usage.html}} \\ \hline
    Installing software & \texttt{nix-shell -p precice} & \texttt{spack install precice \&\& module load ...} & \texttt{eb package.eb -robot \&\& module load ...} \\ \hline
    Installing specific version of a software & \texttt{nix-shell -p boost159} or \texttt{overrideAttrs} & \texttt{spack install gcc@6.4.0} & find old easybuild config and install that one \\ \hline
    Installing software at a point in time & \texttt{nix-shell -p pan -I nixpkgs=\$URL} & \texttt{git checkout sha \&\& spack install precice} & \texttt{pip install} older version  \\ \hline
    Installing software using a specific compiler & \texttt{stdenv} overlay & \texttt{spack install precice \%clang} & \texttt{eb -toolchain=...} \\ \hline
    Setting compiler options & \texttt{stdenv} overlay & \texttt{spack install ... cflags='-O3 -march=native'} & \texttt{\$CFLAGS}, \texttt{\$CXXFLAGS} and respective fortran env vars \\ \hline
  \end{tabular}
\end{figure*}

\subsection{Reproducibility}

\begin{figure}
    \centering
    \includegraphics[width=0.4\textwidth]{figures/nix-tree.png}
    \caption{The dependency graph of the coreutils package, generated by \texttt{nix-store --query --tree \$(nix build nixpkgs\#coreutils --print-out-paths)}}
    \label{fig:nix-tree}
\end{figure}

As previously mentioned, Nix computes an output hash over all its inputs that is unique for this input, so if anything in the inputs changes, such as the version of a dependency, the output hash changes resulting in a different software build.
This guarantees that given the same set of inputs, we are able to compile the same software, including it is dependency tree as you can see in Fig. \ref{fig:nix-tree}.
One thing to mention is, that Nix currently only builds generic architecture build, so the inputs of a derivation do not include any information about the architecture and/or compiler features.
Another thing to mention is, that all software in the \texttt{/nix/store} can only link back to the \texttt{/nix/store} itself.
Nix validates this for every package and breaks links than allowing links back to a file in the home directory or the directory that was used for building the software.
This is required, to guarantee the best possible reproducibly.

Spack does something similar to Nix where it adds hashes to each installation path, but also stores additional files inside the package installation including the complete spec of the package than can later be used to reproducibly build the same package.~\cite{Gamblin_2015}.
It also does something that Nix, currently lacks support for, namely storing the architecture and cpu features that are used in this build.
That way it can provide binary packages that are optimized for specific CPUs.
None the less it has one problem that Nix does not have: it allows packages to link back into system dependencies, for example, while building hpl-2.3 using a gcc 12.2.0 compiled with Spack we see the usage of the following system dependencies \texttt{libm.so}, \texttt{libc.so}, \texttt{libdl.so}, \texttt{libnl-3.so}, \texttt{libnl-route-3.so}, \texttt{librt.so}, \texttt{libutil.so}, \texttt{libpthread.so} as well as the dynamic loader.
Most of these dependencies are stable, and usually do not change on a system update, but it also means that it simply cannot guarantee the best possible reproducibility and it gets worse if the system compiler was used rather than a compiler build from scratch using Spack.
It is also worth mentioning that the Spack developers are aware of this problem and are currently working on building these system libraries themselves~\footnote{\url{https://github.com/spack/spack/issues/39560}}~\footnote{\url{https://github.com/spack/spack/pull/42082}}.

EasyBuild does achieve reproducibility with its toolchains and the approach to keep old package definitions, meaning that they also keep old dependency definitions and each dependency is pinned at a specific version~\cite{Geimer_2014}.
Regarding, CPU architecture and features, EasyBuild does not do generic builds by default but rather prioritizes the idea of optimizing for a specific system, meaning defaulting to \texttt{-march=native} for all packages and it has to manually be disabled~\footnote{\url{https://docs.easybuild.io/controlling-compiler-optimization-flags/\#controlling\_compiler\_optimization\_flags\_optarch\_default}}.
This leads to inherently impure and irreproducible builds as building the same easyconfig on two different host/architectures results in different binaries.
On the other hand, the already mentioned option for \texttt{osdependencies} goes against the reproducibility idea, because any os dependency is not managed by the package manager and might change at any point.
This leads to the problem of binaries linking back to the operating system \texttt{/usr/lib}, that Spack also has.
For instance, a curl build, uses the libraries libssl, libcrypto, libz, libm and libc, all of them being provided by the operating system, and for EasyBuild it includes libraries that change more frequently such as libssl, libcrypto and libz.\\

\subsection{Performance}

With Nix' possibility to override stdenv it is also possible to set custom compiler flags and Nix provides an option \texttt{impureUseNativeOptimizations} to build packages optimized for your hardware, using the \texttt{-march=native} compiler flag, it can then be applied to the package, that is being installed, or every dependency in the dependency graph.~\footnote{\url{https://nixos.wiki/wiki/Build_flags}}
Nix can guarantee these compiler flags for every dependency because Nix wraps its compiler binaries in scripts, meaning it has more control over them and packages can not remove any flags by clearing environment variables.
The advantage of overriding the stdenv is that every package in a dependency tree can be optimized, potentially leading to better overall performance of packages.
But it also has the downside that every package needs to be rebuild potentially taking several days, for example preCICE requires a rebuild of 1500 packages, running for two days on an high-performance computing (HPC) cluster (2x AMD EPYC 7702).
So it might be sufficient to optimize only the package, that is easily possible with \texttt{(package.override)} such as shown before or additionally optimize its direct dependencies, that is possible but more time-consuming to do, because one has to manually override each dependency.

Spack also allows to override cflags, using \texttt{spack install \ldots cflags='-march=native'}, so the package is then being optimized.
By setting cflags with \texttt{==}, the cflags are also applied to the dependencies, causing them to be rebuilt.\footnote{\url{https://spack.readthedocs.io/en/latest/basic_usage.html\#compiler-flags}}
Additionally, there is a way to define flags for a specific registered compiler, but these are only respected by CMake and autotools packages.
Also, a package might also have unavoidable system dependencies and are not optimized, potentially impacting performance negatively.

EasyBuild applies \texttt{-march=native} by default, but you can further customize your compiler flags using \texttt{\$CFLAGS}, \texttt{\$CXXFLAGS}, \texttt{\$FFLAGS}, \texttt{\$F90FLAGS}.
It also has the same problem as Spack that it might have unavoidable unoptimized system package that might impact performance negatively.\\

\subsection{Testing}

We also shortly look at the capabilities of the package managers to run test and evaluate the built packages.
Nix has something that is called a checkPhase that can be run for each package after being built.
There are sane defaults for all available build systems, such as \texttt{make test} or \texttt{ctest} for makefiles and CMake, \texttt{cargo check} and \texttt{cargo test} for rust packages, \texttt{go test} for go packages and \texttt{pytest} for python packages and more.
By default, this checkPhase is not enabled for every build system, such as the \texttt{stdenv} build system, that is responsible for make, CMake and similar, does not enable it but the \texttt{go}, \texttt{rust} and \texttt{python} build system does.
For these build systems, package maintainers have to manually disable the check phase or they can override the check phase if the tests are not working out of the box and make the needed adjustments to resolve the issues.

Spack has something similar, but its not enabled by default, and users, not package maintainers, has to manually specify that they want to run test on installing the package with \texttt{--test=root}.

EasyBuild on the other hand does not offer this feature but has another feature called sanity checks, where EasyBuild ensures that paths and or binaries, specified by a package maintainer inside the package definition, are successfully installed.

Additionally, Nix offers a way for testing multiple packages in combination and or with a specific network setup by providing a way to write your own tests.
These tests are then running inside isolated qemu virtual machines and users are able to start as many VMs as they need and define the set of commands that need to run on each of the so-called nodes.
The tests are pythons scripts, and it is possible to wait until systemd targets are running, until ports are available or wait until commands are completed and compare the output and status codes.

\subsection{Conclusion}

To summarize the comparison, while Nix has the biggest software library, it lacks some scientific software compared to Spack and EasyBuild.
On the other hand, the build sandbox that Nix offers, guarantees a level of reproducibility that Spack and EasyBuild currently cannot provide, because it disallows random linking.
Spack and EasyBuild on the other side both provide a way better usability to end user with their command line interfaces, Nix still is able to do the same things, such as changing the compiler but generally requires knowledge of the Nix language to make this achievable, but if end user have this knowledge it is possible to change the compile flags for all packages in the dependency graph allowing for additional performance gains, if any package is compiled against the specific hardware.

>>>>

All contents, including source files and resulting build artifacts, are stored inside the Nix store.
Per default, the Nix store resides in the path \texttt{/nix/store} on the file system.
The naming scheme for the preCICE 2.5.0 package as an example is \texttt{/nix/store/0a5gw3l\ldots-precice-2.5.0}.
This eases checking the Nix store for integrity and ensures that builds using the same inputs are performed only once.

@@
